### 2019-10-15 ch1_introduction/content.tex
update

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.8\linewidth]{fig/rnaworld} % TODO: change image
	\caption{
		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life. Image from Horning (2011)\cite{horning_rnaworld_2011}.
%		\fbox{\color{red} TODO: combine images, rewrite the description.}
	}
	\label{fig:rnaworld}
\end{figure}

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.7\linewidth]{fig/central_dogma} % TODO: change image
	\caption{
		\textbf{Central Dogma.}
%		\fbox{\color{red} TODO: combine images, rewrite the description.}
	}
	\label{fig:centraldogma}
\end{figure}

@@

\begin{figure}[htb!]
	\centering\includegraphics[width=.8\linewidth]{fig/control_gene_expression} % TODO: change image
	\caption{
		\textbf{Levels of controlling gene expression} can happen at the level of transcription, RNA processing (i.e. splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity \cite{alberts_overviewgenecontrol_2002}.
	}
	\label{fig:controlexpression}
\end{figure}

@@

\begin{itemize}
	\item We \textbf{develop benchmarking strategies} for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}). \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
	\item We apply this strategy to \textbf{perform a comparison of TI methods} (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking. We provide a set of guidelines for end-users wishing to infer trajectories. Our pipeline, datasets, metrics, and containerised wrappers of TI methods are made publicly available for developers to use.
	\item We develop \textbf{a toolkit to infer, visualise and interpret single-cell trajectories} using more than 50 different TI methods (Chapter~\ref{chap:dyno}). dyno provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the tragectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g. adding directionality or adding annotation).
	\item We introduce a TI method specialised in \textbf{inferring linear trajectories} (Chapter~\ref{chap:scorpius}). Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
	\item We develop an algorithm for  \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}). We demonstrate our NI methodology by inferring regulatory interactions for 22'122 hematopoietic cells, as well as for 14'963 bulk profiles of cancer tumours.
	\item We provide a tool for  \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
	\item We reflect on \textbf{challenges in performing self-assessments of TI methods} (Chapter~\ref{chap:selfassessment}).
	Reproducability problems of TI methods are partially due to low rates of quantitative self-assessment. We look at 75 manuscripts of TI methods and identify key challenges the authors had in benchmarking their tool. For each of these challenges, we provide a summary of how we and others have successfully tackled these challenges, in order to spur developers to perform more self-assessments.
	\item Finally, we \textbf{summarise our experience in benchmarking computational methods} by providing a list of essential guidelines in benchmarking computational tools (Chapter~\ref{chap:guidelines}).
\end{itemize}

@@

\begin{table}[ht!]
	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the relation with respect to the package: \textit{aut} Author, \textit{ctb} Contributor. Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-09-10. CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics. In addition, many of the dynverse packages have only recently been published on CRAN. For Github repositories, no download statistics could be retrieved. } \label{tab:packages}

>>>>

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.8\linewidth]{fig/rnaworld}
	\caption{
		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life. Adapted from Horning (2011)\cite{horning_rnaworld_2011}.
	}
	\label{fig:rnaworld}
\end{figure}

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.8\linewidth]{fig/central_dogma.pdf}
	\caption{	
		\textbf{The central dogma of molecular biology explains the flow of genetic information, from DNA to RNA, to protein.}
	}
	\label{fig:centraldogma}
\end{figure}

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\linewidth]{fig/control_gene_expression}
	\caption{
		\textbf{Levels of controlling gene expression} can happen at the level of transcription, RNA processing (i.e. splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity\cite{alberts_overviewgenecontrol_2002}.
	}
	\label{fig:controlexpression}
\end{figure}

@@

\begin{itemize}
	\item We \textbf{develop benchmarking strategies} for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}). \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
	\item We apply this strategy to \textbf{perform a comparison of TI methods} (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking. We provide a set of guidelines for end-users wishing to infer trajectories. Our pipeline, datasets, metrics, and containerised wrappers of TI methods are made publicly available for developers to use.
	\item We develop \textbf{a toolkit to infer, visualise and interpret single-cell trajectories} using more than 50 different TI methods (Chapter~\ref{chap:dyno}). dyno provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the tragectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g. adding directionality or adding annotation).
	\item We introduce a TI method specialised in \textbf{inferring linear trajectories} (Chapter~\ref{chap:scorpius}). Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
	\item We explore a novel algorithm for \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}). We demonstrate our NI methodology by inferring regulatory interactions for 22'122 hematopoietic cells, as well as for 14'963 bulk profiles of cancer tumours.
	\item We provide a tool for \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
	\item We reflect on \textbf{challenges in performing self-assessments of TI methods} (Chapter~\ref{chap:selfassessment}).
	Reproducability problems of TI methods are partially due to low rates of quantitative self-assessment. We look at 75 manuscripts of TI methods and identify key challenges the authors had in benchmarking their tool. For each of these challenges, we provide a summary of how we and others have successfully tackled these challenges, in order to spur developers to perform more self-assessments.
	\item Finally, we \textbf{summarise our experience in benchmarking computational methods} by providing a list of essential guidelines in benchmarking computational tools (Chapter~\ref{chap:guidelines}).
\end{itemize}

@@

\begin{table}[ht!]
	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the relation with respect to the package: \textit{aut} Author, \textit{ctb} Contributor. Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-10-14. CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics. In addition, many of the dynverse packages have only recently been published on CRAN. For Github repositories, no download statistics could be retrieved. } \label{tab:packages}

### 2019-10-15 ch3_dynbenchmark/content.tex
update

\begin{itemize}
	\item \textbf{dyngen}: simulations of gene regulatory networks, available at \href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}
	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}
	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulos_prossttprobabilisticsimulation_2018}
	\item \textbf{Splatter}: simulations of non-linear paths between different expression states \cite{zappia_splattersimulationsinglecell_2017}
\end{itemize}

@@

For more simplistic data generation ("toy" datasets), we created the dyntoy workflow (\url{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}) . We created 12 topology generators (described below), and with 10 datasets per generator, this lead to a total of 120 datasets.

@@

All metrics described here were implemented within the \href{https://github.com/dynverse/dyneval}{dyneval R package (https://github.com/dynverse/dyneval)}.

>>>>

\begin{itemize}
	\item \textbf{dyngen}: simulations of gene regulatory networks, available at \url{https://github.com/dynverse/dyngen}
	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \url{https://github.com/dynverse/dyntoy}
	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulos_prossttprobabilisticsimulation_2018}
	\item \textbf{Splatter}: simulations of non-linear paths between different expression states \cite{zappia_splattersimulationsinglecell_2017}
\end{itemize}

@@

For more simplistic data generation ("toy" datasets), we created the dyntoy workflow (\url{https://github.com/dynverse/dyntoy}). We created 12 topology generators (described below), and with 10 datasets per generator, this lead to a total of 120 datasets.

@@

All metrics described here were implemented within the dyneval R package (\url{https://github.com/dynverse/dyneval}).

### 2019-10-15 ch8_discussion/content.tex
update

\section{Inferring single cell regulatory networks}
\mycomment{TODO: fill in when chapter has been written.}
%	\item We develop an algorithm for  \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}). We demonstrate our NI methodology by inferring regulatory interactions for 25'000 hematopoietic cells, as well as for 15'000 cancer-related bulk profiles. % TODO: was this chapter written in the end?

%\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>^

@@

\textbf{Git}\cite{torvalds_gitfastversion_2005} is a code-revision system that allows multiple users to collaborate on developing code and keep track of what changes were made by whom. 
Since Wouter and I were often working closely together on a specific part of our software, Git saved us a lot of time by merging together changes developed in parallel. Only in few cases did we need to intervene manually to merge our changes.
Additionally, Github.com allowed us not only to collaborate with each-other, but also correspond and collaborate with other software developers from many different research groups, using Github Issues to discuss problems with other researchers, and Github Pull Requests to contribute code to open-source projects. 
% Misschien ook Wouter zijn achternaam gebruiken. Klinkt wat professioneler :-p
Together, we published over 20'000 code contributions (commits) across 50+ software packages\cite{cannoodt_developmentdynverse_2019}. Since many of these packages depend on one another, it is inevitable that changes made to one package would break another package. We wrote many, many unit tests (and still too few), and we let these tests automatically be executed on \textbf{Travis CI}\cite{traviscigmbh_traviscitest_2011}. This way, when we pushed breaking changes to Github, Travis CI would notify that our commit has resulted in one or more unit tests failing. While sometimes it is a hassle to set up, Travis CI has prevented us countless times from generating faulty results due to a faulty underlying function.

>>>>

\section{Inferring single cell regulatory networks}

We proposed a novel algorithm for inferring case-wise gene regulatory interactions. While we demonstrated the methodology by inferring 22'122 cell-specific GRNs and 14'963 sample-specific GRNs from the Tabula Muris and TCGA projects respectively, further interpretation of these results needs to be perform to further validate the usefulness of the method.

Additional avenues for exploration could involve applying trajectory inference on the case-wise regulome data, and quantitatively evaluating the performance of the algorithm using the \texttt{dyngen} simulations.

@@

\textbf{Git}\cite{torvalds_gitfastversion_2005} is a code-revision system that allows multiple users to collaborate on developing code and keep track of what changes were made by whom. 
Since Wouter Saelens and I were often working closely together on a specific part of our software, Git saved us a lot of time by merging together changes developed in parallel. Only in few cases did we need to intervene manually to merge our changes.
Additionally, Github.com allowed us not only to collaborate with each-other, but also correspond and collaborate with other software developers from many different research groups, using Github Issues to discuss problems with other researchers, and Github Pull Requests to contribute code to open-source projects.

Together, we published over 20'000 code contributions (commits) across 50+ software packages\cite{cannoodt_developmentdynverse_2019}. Since many of these packages depend on one another, it is inevitable that changes made to one package would break another package. We wrote many, many unit tests (and still too few), and we let these tests automatically be executed on \textbf{Travis CI}\cite{traviscigmbh_traviscitest_2011}. This way, when we pushed breaking changes to Github, Travis CI would notify that our commit has resulted in one or more unit tests failing. While sometimes it is a hassle to set up, Travis CI has prevented us countless times from generating faulty results due to a faulty underlying function.
